Visual diagnostics for constrained optimisation with application to guided tours

### 2020-12-13 paper/paper.Rmd
modify as per Nicolas suggestion and author

---
title: "Visual diagnostics for constrained optimisation with application to guided tours"
author:
  - name: H.Sherry Zhang
    affiliation: Department of Econometrics and Business Statistics, Monash University 
    email:  Huize.Zhang@monash.edu
abstract: >
  Guided tour searches for interesting low-dimensional views of high-dimensional data via optimising a projection pursuit index function. The first paper of projection pursuit by @friedman1974projection stated that "the  technique used for maximising the projection index strongly influences both the statistical and the computational aspects of the procedure." While many indices have been proposed in the literature, less work has been done on evaluating the performance of the optimisers. In this paper, we implement a data collection object for the optimisation in the guided tour and introduce visual diagnostics based on the data object collected. These diagnostics and workflows can be applied to a broad class of optimisers, to assess their performance. An R package, \pkg{ferrn}, has been created to implement the diagnostics.
preamble: >
  \usepackage{amssymb, amsmath, mathtools, dsfont, bbm, array, booktabs}
  \usepackage[ruled,vlined, linesnumbered]{algorithm2e}
output: rticles::rjournal_article
---

@@

Optimization methods can be divided into various classes, such as global optimisation [@kelley1999iterative; @fletcher2013practical], convex optimisation [@boyd2004convex] or stochastic optimisation [@nocedal2006numerical]. In this paper, we consider methods that rely on the objective function derivatives known as gradient methods, as well as on methods known as derivative-free methods where optimization does not rely on the knowledge of the gradient or other derivatives.

In gradient ascent methods the goal is to find the maximum of an objective function while in gradient descent methods the interest is in the minimum. The maximum is searched for by taking steps proportional to the gradient of the function at the evaluation point, that is $x_{k+1} = x_k +\alpha_kS_k$ where $\alpha_k$ is the step size or the learning rate [@fletcher2013practical]. One difficulty is the choice of step size, as inadequate step sizes might prevent the optimisation method to converge to an extremum. Moreover, these methods are known to be relative slow and can have convergence issues when problems are poorly conditioned [@trefethen1997numerical]. In order to solve some of those problems, line minimization or line search methods were proposed [@shi2004convergence] where the step size is chosen in such a way that convergence is ensured. In line search methods, users are required to provide an initial estimate $x_{1}$ and a search direction $S_k$ for each iteration. Then, the objective is to find the value of $\alpha_k \in \mathbb{R}$ that maximises $f(x_k + \alpha_kS_k)$ with respect to $\alpha_k$ (this step is called the line search). Then one moves on to the next point following $x_{k+1} = x_k + \alpha_kS_k$ and the process is repeated until the desire convergence is reached.
Modern developments of line search methods focus on improving the computation of the searching direction $S_{k}$, and on approximations of the step size $\alpha_k$ together with reducing the computational burden of storing intermediate computations while catering for practical optimisation problems. In the next section, two line search methods for the optimisation used in the projection pursuit guided tour will be presented.

Although gradient optimization methods are very popular, they rely on the availability of the objective function derivatives and on the complexity of the constraints. Because of that, a large class of derivative-free optimization methods have been developed over the years, where the emphasis is on finding, in most cases, a near optimal solution. Examples of those include response surface methodology [@box1951experimental], stochastic approximation [@robbins1951stochastic], random search [@fu2015handbook] and  heuristic methods [@sorensen2013metaheuristics]. Later, we will present a simulated annealing optimisation algorithm, which belongs to the class of random search methods, for optimisation with the guided tour. Moreover, we will introduce an algorithm to further improve optimization solutions with both gradient descent methods and derivative-free optimisation.

Several R implementations address optimization problems with both general purpose as well as task specifics solvers. The most prominent one within the general solvers is `optim()` in the \CRANpkg{stats} [@stats] package, which provides both gradient-based and derivative-free optimisation functions. Another general solver specialised in non-linear optimisation is \CRANpkg{nloptr} [@nloptr]. For more specific optimisation solvers, readers are recommended to visit the relevant sections in CRAN task review on *optimisation and mathematical programming* [@crantaskreviewoptim].

>>>>

---
title: "Visual diagnostics for constrained optimisation with application to guided tours"
author:
  - name: H.Sherry Zhang
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  huize.zhang@monash.edu
  - name: Dianne Cook
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  dicook@monash.edu
  - name: Ursula Laa
    affiliation: University of Natural Resources and Life Sciences
    address: Institute of Statistics
    email:  ursula.laa@boku.ac.at  
  - name: Nicolas Langrené
    affiliation: Data61 CSIRO 
    address: Locked Bag 38004 Docklands VIC 8012 Australia
    email: Nicolas.Langrene@data61.csiro.au
  - name: Patricia Menéndez
    affiliation: Monash University 
    address: Department of Econometrics and Business Statistics
    email:  patricia.menendez@monash.edu

abstract: >
  Guided tour searches for interesting low-dimensional views of high-dimensional data via optimising a projection pursuit index function. The first paper of projection pursuit by @friedman1974projection stated that "the  technique used for maximising the projection index strongly influences both the statistical and the computational aspects of the procedure." While many indices have been proposed in the literature, less work has been done on evaluating the performance of the optimisers. In this paper, we implement a data collection object for the optimisation in the guided tour and introduce visual diagnostics based on the data object collected. These diagnostics and workflows can be applied to a broad class of optimisers, to assess their performance. An R package, \pkg{ferrn}, has been created to implement the diagnostics.
preamble: >
  \usepackage{amssymb, amsmath, mathtools, dsfont, bbm, array, booktabs}
  \usepackage[ruled,vlined, linesnumbered]{algorithm2e}
output: rticles::rjournal_article
---

@@

Optimization methods can be divided into various classes, such as global optimisation [@kelley1999iterative; @fletcher2013practical], convex optimisation [@boyd2004convex] or stochastic optimisation [@nocedal2006numerical]. A large class of methods utilises the gradient information in the objective function with the most notable one being the gradient ascent (descent) method. Although gradient optimization methods are popular, they rely on the availability of the objective function derivatives and on the complexity of the constraints. Because of that, we consider another class of method in this paper: derivative-free methods. Derivative-free methods have been developed over the years, where the emphasis is on finding, in most cases, a near optimal solution. Examples of those include response surface methodology [@box1951experimental], stochastic approximation [@robbins1951stochastic], random search [@fu2015handbook] and heuristic methods [@sorensen2013metaheuristics]. Later, we will present a simulated annealing optimisation algorithm, which belongs to the class of random search methods, for optimisation with the guided tour.

<!-- In this paper, we consider derivative free method without the knowledge of the gradient. Somd df minic the graident method, biggest class is line search, then SA -->

<!--  In the case of gradient ascent, the maximum is searched by taking steps proportional to the search direction at the evaluation point, that is $x_{k+1} = x_k +\alpha_kS_k$ where $\alpha_k$ is the step size or the learning rate [@fletcher2013practical].  One difficulty is the choice of step size, as inadequate step sizes might prevent the optimisation method to converge to an extremum. Moreover, these methods are known to be relative slow and can have convergence issues when problems are poorly conditioned [@trefethen1997numerical]. -->

A common search scheme utilised by both derivative-free methods and gradient methods is line search. In line search methods, users are required to provide an initial estimate $x_{1}$ and at each iteration, a search direction $S_k$ and a step size $\alpha_k$ are generated. Then one moves on to the next point following $x_{k+1} = x_k + \alpha_kS_k$ and the process is repeated until the desire convergence is reached. While gradient-based methods choose the search direction by the gradient, derivative-free methods determine the search direction through local information of the objective function. The choice of step size also needs considerations, as inadequate step sizes might prevent the optimisation method to converge to an extremum. An ideal step size can be chosen via finding the value of $\alpha_k \in \mathbb{R}$ that maximises $f(x_k + \alpha_kS_k)$ with respect to $\alpha_k$ at each iteration.

<!-- Modern developments of line search methods focus on improving the computation of the searching direction $S_{k}$, and on approximations of the step size $\alpha_k$ together with reducing the computational burden of storing intermediate computations while catering for practical optimisation problems. -->

<!-- methods that rely on the objective function derivatives known as gradient methods, as well as on methods known as derivative-free methods where optimization does not rely on the knowledge of the gradient or other derivatives. -->

<!-- In gradient ascent methods the goal is to find the maximum of an objective function while in gradient descent methods the interest is in the minimum.  -->
 <!-- In order to solve some of those problems, line minimization or line search methods were proposed [@shi2004convergence] where the step size is chosen in such a way that convergence is ensured.  -->

 <!--  In the next section, two line search methods for the optimisation used in the projection pursuit guided tour will be presented. -->

Several R implementations address optimization problems with both general purpose as well as task specifics solvers. The most prominent one within the general solvers is `optim()` in the \CRANpkg{stats} [@stats] package, which provides both gradient-based and derivative-free optimisation functions. Another general solver specialised in non-linear optimisation is \CRANpkg{nloptr} [@nloptr]. Specific solvers for simulated annealing includes `optim(..., method  "SANN")` and package \CRANpkg{GenSA} [@gensa] that deals with more complicated objective functions. For other task specific solvers, readers are recommended to visit the relevant sections in CRAN task review on *optimisation and mathematical programming* [@crantaskreviewoptim].

### 2020-12-13 paper/paper.tex
modify as per Nicolas suggestion and author

% !TeX root = RJwrapper.tex
\title{Visual diagnostics for constrained optimisation with application to
guided tours}
\author{by H.Sherry Zhang}

@@

Optimization problems are ubiquitous in many areas of study. While in
some cases, analytical solutions can be found, the majority of problems
rely on numerical methods to find the optimal solution. These numerical
methods follow interactive approaches that aim at finding the optimum by
progressively improving the current solution until a desirable accuracy
is achieved. Although that principle seems uncomplicated, optimization
methods need to deal with many challenges such as the existence of
multiple maxima (local and global), the trade off between desirable
accuracy and computational burden, as well as dealing with noisy
objective functions and possible constraints. In addition, optimization
results might depend on the algorithm starting values affecting the
consistency of results.

Our interest is on constrained optimization
\citep{bertsekas2014constrained} as defined in the introduction section
and assume that it is not possible to find a solution to the problem in
the way of a closed form. That is, the problem consists of finding the
minimum or maximum of a function \(f \in L^p\) in the constrained
\(\mathbb{A}\) space.

Optimization methods can be divided in various classes, such as global
optimisation \citep{kelley1999iterative, fletcher2013practical}, convex
optimisation \citep{boyd2004convex}, stochastic optimisation
\citep{nocedal2006numerical} . In this paper, we focused on methods that
rely on the objective function derivatives known as gradient methods,
and those which are known as derivative free methods where optimization
does not rely on the gradient or derivatives information.

In gradient ascent methods the goal is to find the maximum of an
objective function while in gradient descent methods the interest is in
the minimum. The maximum is searched for by taken steps proportional to
the gradient of the function at the evaluation point that is
\(x_{k+1} = x_k +\alpha_kS_k\) where \(\alpha_k\) is the step size or
the learning rate \citep{fletcher2013practical}. The step size is the
key in the success of the optimization problem as inadequate step sizes
might lead to a failure in convergence. These methods are known to be
relative slow and can have convergence issues when problems are poorly
conditioned \citep{trefethen1997numerical}. In order to solve some of
those problems, line minimization or line search methods were proposed
\citep{shi2004convergence} where the step size is chosen in such a way
that convergence is ensured. In line search methods users are required
to provide an initial estimate \(x_{1}\), a search direction \(S_k\) for
each iteration where the objective is to find the value of
\(\alpha_k \in \mathbb{R}\) that maximises \(f(x_k + \alpha_kS_k)\) with
respect to \(\alpha_k\) (this step is called the line search). Then move
on to the next point following \(x_{k+1} = x_k + \alpha_kS_k\) and
repeat the process until the desire convergence is reached. Modern
development of line search methods focuses on improving the computation
of the searching direction \(S_{k}\), and on approximations of the step
size \(\alpha_k\) together with reducing the computational storing
burden of the intermediate computations while catering for practical
optimisation problems. In the next section two line search methods for
the optimisation used in the projection pursuit guided tour will be
presented.

Although gradient optimization methods are very popular, they rely on
the existence of the objective function derivatives and on the
complexity of the constraints. Because of that, the second big class of
optimization problems look at derivative free methods where the emphasis
is on finding, in most cases, a near optimal solution. Examples of those
include response surface methodology \citep{box1951experimental},
stochastic approximation \citep{robbins1951stochastic}, random search
\citep{fu2015handbook} and heuristic methods
\citep{sorensen2013metaheuristics}. Later, we will present an
optimization algorithm within the random search class, namely simulated
annealing for optimization with the guided tour. Moreover, we will
introduce an algorithm to further improve optimization solutions both in
gradient descent methods and in derive free optimization.

Several R implementations address optimization problems with both
general purpose as well as task specifics solvers. The most prominent
one within the general solvers is \texttt{optim()} in the
\CRANpkg{stats} \citep{stats} package, which have functions for both
gradient based and derivative free methods. Another general solver
specialised in non-linear optimisation is \CRANpkg{nloptr}
\citep{nloptr}. For more specific optimisation solvers, readers are
recommended to svisite the relevant sections in CRAN task review on
\emph{optimisation and mathematical programming}
\citep{crantaskreviewoptim}.

@@

\address{%
H.Sherry Zhang\\
Department of Econometrics and Business Statistics, Monash University\\
\\
}
\href{mailto:Huize.Zhang@monash.edu}{\nolinkurl{Huize.Zhang@monash.edu}}

>>>>

% !TeX root = RJwrapper.tex
\title{Visual diagnostics for constrained optimisation with application to
guided tours}
\author{by H.Sherry Zhang, Dianne Cook, Ursula Laa, Nicolas Langrené, Patricia Menéndez}

@@

Optimization problems are ubiquitous in many areas of study. While in
some cases analytical solutions can be found, the majority of problems
rely on numerical methods to find the optimal solution. These numerical
methods follow iterative approaches that aim at finding the optimum by
progressively improving the current solution until a desirable accuracy
is achieved. Although that principle seems uncomplicated, optimization
methods need to deal with many challenges such as the possible existence
of multiple maxima (local and global), the trade-off between desirable
accuracy and computational burden, as well as dealing with noisy
objective functions and possible constraints. In addition, the
optimization results might depend on the algorithm starting values,
affecting the consistency of results.

Our interest is on constrained optimization
\citep{bertsekas2014constrained} as defined in the introduction section,
and assuming it is not possible to find a solution to the problem in the
way of a closed-form. That is, the problem consists of finding the
minimum or maximum of a function \(f \in L^p\) in the constrained
\(\mathbb{A}\) space.

Optimization methods can be divided into various classes, such as global
optimisation \citep{kelley1999iterative, fletcher2013practical}, convex
optimisation \citep{boyd2004convex} or stochastic optimisation
\citep{nocedal2006numerical}. A large class of methods utilises the
gradient information in the objective function with the most notable one
being the gradient ascent (descent) method. Although gradient
optimization methods are popular, they rely on the availability of the
objective function derivatives and on the complexity of the constraints.
Because of that, we consider another class of method in this paper:
derivative-free methods. Derivative-free methods have been developed
over the years, where the emphasis is on finding, in most cases, a near
optimal solution. Examples of those include response surface methodology
\citep{box1951experimental}, stochastic approximation
\citep{robbins1951stochastic}, random search \citep{fu2015handbook} and
heuristic methods \citep{sorensen2013metaheuristics}. Later, we will
present a simulated annealing optimisation algorithm, which belongs to
the class of random search methods, for optimisation with the guided
tour.

A common search scheme utilised by both derivative-free methods and
gradient methods is line search. In line search methods, users are
required to provide an initial estimate \(x_{1}\) and at each iteration,
a search direction \(S_k\) and a step size \(\alpha_k\) are generated.
Then one moves on to the next point following
\(x_{k+1} = x_k + \alpha_kS_k\) and the process is repeated until the
desire convergence is reached. While gradient-based methods choose the
search direction by the gradient, derivative-free methods determine the
search direction through local information of the objective function.
The choice of step size also needs considerations, as inadequate step
sizes might prevent the optimisation method to converge to an extremum.
An ideal step size can be chosen via finding the value of
\(\alpha_k \in \mathbb{R}\) that maximises \(f(x_k + \alpha_kS_k)\) with
respect to \(\alpha_k\) at each iteration.

Several R implementations address optimization problems with both
general purpose as well as task specifics solvers. The most prominent
one within the general solvers is \texttt{optim()} in the
\CRANpkg{stats} \citep{stats} package, which provides both
gradient-based and derivative-free optimisation functions. Another
general solver specialised in non-linear optimisation is
\CRANpkg{nloptr} \citep{nloptr}. Specific solvers for simulated
annealing includes \texttt{optim(...,\ method\ \ "SANN")} and package
\CRANpkg{GenSA} \citep{gensa} that deals with more complicated objective
functions. For other task specific solvers, readers are recommended to
visit the relevant sections in CRAN task review on \emph{optimisation
and mathematical programming} \citep{crantaskreviewoptim}.

@@

\address{%
H.Sherry Zhang\\
Monash University\\
Department of Econometrics and Business Statistics\\
}
\href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}

\address{%
Dianne Cook\\
Monash University\\
Department of Econometrics and Business Statistics\\
}
\href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}

\address{%
Ursula Laa\\
University of Natural Resources and Life Sciences\\
Institute of Statistics\\
}
\href{mailto:ursula.laa@boku.ac.at}{\nolinkurl{ursula.laa@boku.ac.at}}

\address{%
Nicolas Langrené\\
Data61 CSIRO\\
Locked Bag 38004 Docklands VIC 8012 Australia\\
}
\href{mailto:Nicolas.Langrene@data61.csiro.au}{\nolinkurl{Nicolas.Langrene@data61.csiro.au}}

\address{%
Patricia Menéndez\\
Monash University\\
Department of Econometrics and Business Statistics\\
}
\href{mailto:patricia.menendez@monash.edu}{\nolinkurl{patricia.menendez@monash.edu}}
